---
title: "lab3_whitney_willoz"
author: "Whitney Willoz"
date: "9/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 3

```{r}
dat1 <- readxl::read_xls('GSS.xls')
library(ggplot2)
library(dplyr)

dat <- dat1 %>%
  rename(Year = `Gss year for this respondent`, Happy = `General happiness`,
         relig = `Rs religious preference`, party = `Political party affiliation`,
         gender = `Respondents sex`, edu = `Highest year of school completed`,
         marStatus = `Marital status`, hoursWrk = `Number of hours usually work a week`, id = `Respondent id number`, Ballot = `Ballot used for interview`) 

```


1) Find the average years of education for the respondents with each marital status. Arrange your output in a meaningful order and print. Describe any patterns you find.

Answer:The highest average for years of education comes from Never married at 13.4 and lowest average is widowed at 11.2 years.

```{r}
MarriageAvgEdu <- dat %>% 
  filter(nchar(edu) < 3) %>% 
  na.omit(edu, marStatus) %>% 
  group_by(marStatus)%>%
  summarize(edu_mean = mean(as.numeric(edu), na.rm=TRUE)) 
MarriageAvgEdu
```

2)Create a single data frame containing records for the better educated respondents with each marital status. A “better educated” respondent is someone who has strictly more years of education than the average among those with the same marital status. Print the structure of the data frame.

```{r}
btrEdu <- dat %>%
  filter((marStatus == "Never married" & edu > 13.4)|
        (marStatus == "Married" & edu > 13)|
        (marStatus == "Divorced" & edu > 13.1)|
        (marStatus == "Widowed" & edu > 11.2)|
        (marStatus== "Separated" & edu > 11.9))
str(btrEdu)
```

For each of the investigation below, create at least one numerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.

3)How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

Answer: Overall there are more married people who are happy than all other categories. Never being married also has a higher proportion.On average Married people are happy 70% of the time. 

```{r}
dat$happy1 <- 
  case_when(
    dat$Happy %in% c('Pretty happy', 'Very happy') ~ 'Happy',
    dat$Happy == 'Not too happy' ~ 'Sad',
    TRUE ~ 'Unknown'
  )
table(dat$happy1)

dat$happy2 <- 
  case_when(
    dat$Happy %in% c('Pretty happy', 'Very happy') ~ 1,
    dat$Happy == 'Not too happy' ~ -1,
    TRUE ~ 0
  )
meanHappytoMar <- dat %>%
  group_by(marStatus)%>%
  summarize(happy_avg = mean(happy2))
meanHappytoMar

ggplot(dat, aes(marStatus, fill = happy1)) + geom_bar()

```

4)Does party affiliation affect the relationship you found in the last question?

Answer:Overall, married not strong democrats and republicans are overall the happiest or the highest in volume.Next would be married Strong democrats or independents.

```{r}
meanHappywParty <- dat %>%
  group_by(marStatus, party)%>%
  summarize(happy_avg = mean(happy2))
meanHappywParty
ggplot(dat, aes(marStatus, fill=happy1)) + geom_bar() + coord_flip() + facet_wrap(~party)
```

5)How is religion related to education? 

Answer: People who practice hinduism have the highest average of years of education followed by jewish.

```{r}
avgEduByRelig <- dat %>%
  group_by(relig) %>%
  filter(nchar(edu) < 3)%>%
  summarize(edu_mean = mean(as.numeric(edu), na.rm = TRUE))
avgEduByRelig

 dat %>%
  group_by(relig) %>%
  filter(nchar(edu) < 3)%>%
  summarize(edu_mean = mean(as.numeric(edu), na.rm = TRUE)) %>% 
  ggplot(aes(relig,edu_mean)) + geom_bar(stat = 'identity') + coord_flip()
```