---
title: "lab3_whitney_willoz"
author: "Whitney Willoz"
date: "9/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 3

```{r}
install.packages('readxl')
dat <- readxl::read_xls('C:\Users\wrwil\OneDrive\Documents\202rFiles\extract1.tar/GSS.xls')
library(ggplot2)
library(dplyr)
```


1) Find the average years of education for the respondents with each marital status. Arrange your output in a meaningful order and print. Describe any patterns you find.

Answer:

```{r}
singleAvgEdu <- dat %>% 
  select(marStatus, edu) %>% 
  filter(marStatus == "Single") %>% 
  mean(edu)
singleAvgEdu

marriedAvgEdu <- dat %>% 
  select(marStatus, edu) %>% 
  filter(marStatus == "Single") %>% 
  mean(edu)
marriedAvgEdu
```

2)Create a single data frame containing records for the better educated respondents with each marital status. A “better educated” respondent is someone who has strictly more years of education than the average among those with the same marital status. Print the structure of the data frame.

```{r}
btrEdu <- dat %>%
  filter()
```

For each of the investigation below, create at least one numerical summary and one appropriate visualization. Write a paragraph to describe your findings and the variables you used.

3)How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.

Answer:

```{r}
happyRating <- dat %>%
  select(happiness, marStatus) 
happinessBar <- ggplot(happyRating, aes(marStatus, fill = happiness) ) + geom_bar()
happinessBar

```

4)Does party affiliation affect the relationship you found in the last question?

Answer:

```{r}

```

5)How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?

Answer: 

```{r}
singleAvgEdu <- dat %>% 
  select(marStatus, edu) %>% 
  filter(marStatus == "Single") %>% 
  mean(edu)
singleAvgEdu

marriedAvgEdu <- dat %>% 
  select(marStatus, edu) %>% 
  filter(marStatus == "Single") %>% 
  mean(edu)
marriedAvgEdu

ggplot(dat, aes()) + geom_boxplot()
```